# Reproducible Research: Peer Assessment 1

## Loading and preprocessing the data

```{r eval=-2}
library(lattice)
system("unzip activity.zip")
data1 <- read.csv("activity.csv")
summary(data1)
```
```{r}
# Function to print numbers to decimal places by default
print2 <- function(x, decimals=2) noquote(format(x, nsmall=decimals))

# function to return the direction and magnitude of a change as a string
strChange <- function(x) {
    if (x == 0) return("no effect")
    if (x > 0) return(paste("increase of", x))
    if (x < 0) return(paste("decrease of", abs(x)))
}
```

## What is mean total number of steps taken per day?
Compute total number of steps taken each day:
```{r}
totDailySteps <- with(data1, by(steps, date, sum))
```
```{r}
hist(totDailySteps, breaks=7)
```

Compute mean daily steps:
```{r}
print2(meanDailyStepsRaw <- mean(totDailySteps, na.rm=TRUE))
```
Compute median daily steps:
```{r}
print2(medianDailyStepsRaw <- as.numeric(median(totDailySteps, na.rm=TRUE)))
```

## What is the average daily activity pattern?
Compute average number of steps taken for each 5-minute interval:
```{r}
avgStepsPerInterval <- with(data1, by(steps, interval, mean, na.rm=TRUE))
```

Plot the average number of steps per 5-min. interval by the 5-min. interval.

```{r fig.height=4, fig.width=8}
xvar <- as.numeric(names(avgStepsPerInterval)) 
yvar <- avgStepsPerInterval
plot(xvar, yvar, type='l', xlab="5-minute inteval", ylab="Average number of steps")
```

Determine which interval has the highest average number of steps:

```{r}
print2(maxSteps <- max(avgStepsPerInterval))
print2(maxStepInterval = names(which(avgStepsPerInterval == maxSteps)))
```
The maximum average number of steps per interval is **`r maxSteps`** and occurs in interval **`r maxStepInterval`**.

## Imputing missing values

The results of `summary data1` at the top of the page show that that `steps` is
the only variable with missing values.  This is verified below:

```{r}
apply(data1, 2, anyNA)  # TRUE only for variables with at least 1 NA
```

```{r} 
# Number of missing values for *steps* variable
stepsNAs <- is.na(data1$steps)
(countOfStepsNAs <- length(stepsNAs[stepsNAs]))
```

Thus there are **`r countOfStepsNAs`** missing values in the dataset, all occurring
in the variable `steps`.

We will impute missing values of `steps` for an `interval` by replacing the
missing value with the mean of the non-missing `steps` for that interval accross
all days in the dataset.

```{r}
# Get a vector of length==nrow(data1) whose values are the mean number of steps
# for each inverval
aveSteps <- with(data1, ave(steps, interval, FUN=function(x) mean(x, na.rm=TRUE)))

# Duplicate data1
data2 <- data1

# Replace missing values in new dataset
data2$steps[is.na(data2$steps)] <- aveSteps[is.na(data2$steps)]

summary(data2)
```
Compute total number of steps taken each day in the imputed dataset:
```{r}
totDailySteps2 <- with(data2, by(steps, date, sum))
```
```{r}
hist(totDailySteps2, breaks=7)
```

Effect of missing value impuation on mean and median number of steps per day:
```{r R.options="digits=9"}
# Mean before imputation
print2(meanDailyStepsRaw)

# Mean after imputation
meanDailyStepsImputed <- mean(totDailySteps2)
print2(meanDailyStepsImputed)

# Difference
print2(dMean <- meanDailyStepsImputed - meanDailyStepsRaw)

# Median before imputation
print2(medianDailyStepsRaw)

# Median after imputation
medianDailyStepsImputed <- median(totDailySteps2)
names(medianDailyStepsImputed) <- NULL
print2(medianDailyStepsImputed)

# Difference
(dMedian <- round(medianDailyStepsImputed - medianDailyStepsRaw, 2))
```

Imputing missing values for `steps` had the following effects on the mean and
median number of steps per day:
* Mean: `r strChange(dMean)`
* Median: `r strChange(dMedian)`

## Are there differences in activity patterns between weekdays and weekends?

```{r}
# Create factor "weekday" with levels "weekday" and "weekend"
day <- weekdays(as.POSIXct(data2$date), abbreviate=TRUE)
data2$weekday <- factor(ifelse(day %in% c("Sat", "Sun"), "weekend", "weekday"))

# Compute average steps per interval for weekday and weekend data
wkday.avgs <- with(data2[data2$weekday=="weekday",], by(steps, interval, mean))
wkend.avgs <- with(data2[data2$weekday=="weekend",], by(steps, interval, mean))

# Create a dataset for the average step data
avg.data <- data.frame(steps=c(wkday.avgs, wkend.avgs), 
                       interval=as.numeric(c(names(wkday.avgs), names(wkend.avgs))),
                       weekday=c(rep.int("weekday", length(wkday.avgs)),
                                 rep.int("weekend", length(wkend.avgs))))

xyplot(steps ~ interval | weekday, data=avg.data, type='l', layout=c(1,2), xlab="Interval",
       ylab="Average number of steps")
```
