---
output: html_document
---
# Reproducible Research: Peer Assessment 1

## Loading and preprocessing the data

```{r}
library(lattice)
source("formatters.R")  # Functions for pretty printing
unzip("activity.zip")
data1 <- read.csv("activity.csv")
summary(data1)
```
## What is mean total number of steps taken per day?
Compute total number of steps taken each day:
```{r}
totDailySteps <- with(data1, by(steps, date, sum))
```
```{r fig.height=5}
hist(totDailySteps, breaks=7)
```

Compute mean daily steps:
```{r}
print2(meanDailyStepsRaw <- mean(totDailySteps, na.rm=TRUE))
```
Compute median daily steps:
```{r}
print2(medianDailyStepsRaw <- as.numeric(median(totDailySteps, na.rm=TRUE)))
```

## What is the average daily activity pattern?
Compute average number of steps taken for each 5-minute interval:
```{r}
avgStepsPerInterval <- with(data1, tapply(steps, interval, mean, na.rm=TRUE))
```

Plot the average number of steps per 5-min. interval by the 5-min. interval.

```{r fig.height=4, fig.width=12, fig.height=6}
plot(0:287, avgStepsPerInterval, type='l', xlab='', xaxt='n',
     ylab="Average number of steps in 5-min interval", 
     cex.lab=1.2)
xtics <- seq(0, 288, by=12)
xtics2 <- seq(6, 287, by=12)
xlabs <- sprintf("%04d", 100 * xtics / 12)
xaxis <- axis(side=1, at=xtics, labels=xlabs, las=3, tck=-.034)
xaxis2 <- axis(side=1, at=xtics2, labels=F, tck=-.025)
rug(1:287, ticksize=-.017)
mtext("5-min interval (24-hr time)", side=1, line=4, cex=1.2)
title("Average Daily Number of Steps Per 5-min Interval")
```

Determine which interval has the highest average number of steps:

```{r}
# Highest average number of steps in any interval
print2(maxSteps <- max(avgStepsPerInterval))

# Interval with the highest average number of steps
print2(maxStepInterval <- names(which(avgStepsPerInterval == maxSteps)))
```
The maximum average number of steps per interval is **`r print2(maxSteps)`** and occurs 
in interval **`r sprintf("%04d", as.numeric(maxStepInterval))`**.

## Imputing missing values

The results of `summary data1` at the top of the page show that that `steps` is
the only variable with missing values.  This is verified below:

```{r}
apply(data1, 2, anyNA)  # TRUE only for variables with at least 1 NA
```

```{r} 
# Number of missing values for *steps* variable
stepsNAs <- is.na(data1$steps)
(countOfStepsNAs <- length(stepsNAs[stepsNAs]))
```

Thus there are **`r countOfStepsNAs`** missing values in the dataset, all occurring
in the variable `steps`.

We will impute missing values of `steps` for an `interval` by replacing the
missing value with the mean of the non-missing `steps` for that interval accross
all days in the dataset.

```{r}
# Get a vector of length==nrow(data1) whose values are the mean number of steps
# for each inverval
aveSteps <- with(data1, ave(steps, interval, FUN=function(x) mean(x, na.rm=TRUE)))

# Duplicate data1
data2 <- data1

# Replace missing values in new dataset
data2$steps[is.na(data2$steps)] <- aveSteps[is.na(data2$steps)]

summary(data2)
```
Compute total number of steps taken each day in the imputed dataset:
```{r}
totDailySteps2 <- with(data2, by(steps, date, sum))
```
```{r fig.height=5}
hist(totDailySteps2, breaks=7)
```

Effect of missing value impuation on mean and median number of steps per day:
```{r R.options="digits=9"}
# Mean before imputation
print2(meanDailyStepsRaw)

# Mean after imputation
meanDailyStepsImputed <- mean(totDailySteps2)
print2(meanDailyStepsImputed)

# Difference
print2(dMean <- meanDailyStepsImputed - meanDailyStepsRaw)

# Median before imputation
print2(medianDailyStepsRaw)

# Median after imputation
medianDailyStepsImputed <- median(totDailySteps2)
names(medianDailyStepsImputed) <- NULL
print2(medianDailyStepsImputed)

# Difference
(dMedian <- round(medianDailyStepsImputed - medianDailyStepsRaw, 2))
```

Imputing missing values for `steps` had the following effects on the mean and
median number of steps per day:
* Mean: `r strChange(dMean)`
* Median: `r strChange(dMedian)`

## Are there differences in activity patterns between weekdays and weekends?

```{r fig.width=12, fig.height=8}
# Create factor "weekday" with levels "weekday" and "weekend"
day <- weekdays(as.POSIXct(data2$date), abbreviate=TRUE)
data2$weekday <- factor(ifelse(day %in% c("Sat", "Sun"), "weekend", "weekday"))

# Create a dataset for the average step data
avg.data <- data.frame(with(data2, aggregate(steps, list(interval, weekday), mean)))
names(avg.data) <- c('interval', 'weekday', 'steps')

# Plot the average number of steps per day for each interval, averagesd separately 
# over all weekdays and all weekend days, vs. the interval
avg.data$x <- 0:287
xticks3 <- seq(0, 288, by=6)

xlabs3 <- numeric(49)
xlabs3[1:49 %% 2 == 1] <- as.character(sprintf("%04d", xtics*100/12))
xlabs3[1:49 %% 2 == 0] <- ''
xyplot(steps ~ x | weekday, data=avg.data, type='l', layout=c(1,2),
       xlab=list("5-min Interval (24-hr time)", cex=1.2), 
       ylab=list("Average daily number of steps", cex=1.2), 
       scales=list(cex=1.05, x=list(at=xticks3, rot=90, labels=xlabs3)),
       par.strip.text=list(cex=1.15))
```

